---
title: "Table 2"
author: "Gina Guzzo"
date: "17/01/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Load libraries and packages.

```{r, include = FALSE}
library(kableExtra)
library(rmarkdown)
set.seed(42)
library(qwraps2)
# define the markup language we are working in.
# options(qwraps2_markup = "latex") is also supported.
options(qwraps2_markup = "markdown")
library(readxl) 
library(dplyr)
library(tidyr)

```

## 1. Data cleaning
Import data. 
Split cost columns that have more than one value into sub-categories, using commas as separators.

```{r, include = FALSE}
metadata <- read_excel("all_metadata_financial.xlsx")

metadata <- metadata %>%
  separate(`exam (qty, cost)`, into = c("exam_qty", "exam_cost"), sep = ", ", convert = TRUE) %>%
  separate(`bw (qty, cost)`, into = c("bw_qty", "bw_cost"), sep = ", ", convert = TRUE) %>%
  separate(`opg (qty, cost)`, into = c("opg_qty", "opg_cost"), sep = ", ", convert = TRUE) %>%
  separate(`fluor (qty, cost)`, into = c("fluor_qty", "fluor_cost"), sep = ", ", convert = TRUE) %>%
  separate(`scrp (qty, cost)`, into = c("scrp_qty", "scrp_cost"), sep = ", ", convert = TRUE) %>%
  separate(`perio (qty, cost)`, into = c("perio_qty", "perio_cost"), sep = ", ", convert = TRUE) %>%
  separate(`sealants (est qty, est cost, act qty, act cost)`, into = c("seal_est_qty", "seal_est_cost", "seal_act_qty", "seal_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`5_surf_resto (est qty, est cost, act qty, act cost)`, into = c("5_surf_resto_est_qty", "5_surf_resto_est_cost", "5_surf_resto_act_qty", "5_surf_resto_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`4_surf_resto (est qty, est cost, act qty, act cost)`, into = c("4_surf_resto_est_qty", "4_surf_resto_est_cost", "4_surf_resto_act_qty", "4_surf_resto_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`3_surf_resto (est qty, est cost, act qty, act cost)`, into = c("3_surf_resto_est_qty", "3_surf_resto_est_cost", "3_surf_resto_act_qty", "3_surf_resto_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`2_surf_resto (est qty, est cost, act qty, act cost)`, into = c("2_surf_resto_est_qty", "2_surf_resto_est_cost", "2_surf_resto_act_qty", "2_surf_resto_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`1_surf_resto (est qty, est cost, act qty, act cost)`, into = c("1_surf_resto_est_qty", "1_surf_resto_est_cost", "1_surf_resto_act_qty", "1_surf_resto_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`resto_temp (qty, cost)`, into = c("resto_tmp_qty", "resto_tmp_cost"), sep = ", ", convert = TRUE) %>%
  separate(`resto_other (qty, cost)`, into = c("resto_oth_qty", "resto_oth_cost"), sep = ", ", convert = TRUE) %>%
  separate(`exo (est qty, est cost, act qty, act cost)`, into = c("exo_est_qty", "exo_est_cost", "exo_act_qty", "exo_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`endo_rct (qty, cost)`, into = c("endo_rct_qty", "endo_rct_cost"), sep = ", ", convert = TRUE) %>%
  separate(`surg_other (qty, cost)`, into = c("surg_oth_qty", "surg_oth_cost"), sep = ", ", convert = TRUE) %>%
  separate(`occ_splint (est qty, est cost, act qty, act cost)`, into = c("occ_splint_est_qty", "occ_splint_est_cost", "occ_splint_act_qty", "occ_splint_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`ohi (qty, cost)`, into = c("ohi_qty", "ohi_cost"), sep = ", ", convert = TRUE) %>%
  separate(`odont (qty, cost)`, into = c("odont_qty", "odont_cost"), sep = ", ", convert = TRUE)
```


Clean up columns with costs to remove $ symbol.
```{r}
cost_cols <- c(
  "exam_cost", "bw_cost", "opg_cost", "fluor_cost", "scrp_cost", "perio_cost",
  "seal_est_cost", "seal_act_cost",
  "5_surf_resto_est_cost", "5_surf_resto_act_cost",
  "4_surf_resto_est_cost", "4_surf_resto_act_cost",
  "3_surf_resto_est_cost", "3_surf_resto_act_cost",
  "2_surf_resto_est_cost", "2_surf_resto_act_cost",
  "1_surf_resto_est_cost", "1_surf_resto_act_cost",
  "resto_tmp_cost", "resto_oth_cost",
  "exo_est_cost", "exo_act_cost",
  "endo_rct_cost", "surg_oth_cost",
  "occ_splint_est_cost", "occ_splint_act_cost",
  "ohi_cost", "odont_cost",
  "prosth (cost)"
)

metadata <- metadata %>%
  mutate(across(
    all_of(cost_cols),
    ~ as.numeric(gsub("\\$", "", .x))
  ))
```


Make broader categories for treatment types by summing columns.
```{r}
metadata <- metadata %>%
  mutate(
    # Sum diagnostic costs
    diagnostics_cost = rowSums(select(., exam_cost, bw_cost, opg_cost), na.rm = TRUE),
    
    # Sum preventive treatment costs
    prevent_tmnt_cost = rowSums(select(., 
      fluor_cost, scrp_cost, seal_act_cost, ohi_cost, odont_cost, occ_splint_act_cost), na.rm = TRUE),
    
    # Sum restoration costs
    restorations_cost = rowSums(select(., 
      `5_surf_resto_act_cost`, `4_surf_resto_act_cost`, 
      `3_surf_resto_act_cost`, `2_surf_resto_act_cost`, 
      `1_surf_resto_act_cost`, resto_tmp_cost, resto_oth_cost), na.rm = TRUE),
    
    # Sum prosthodontic costs
    prosth_cost = `prosth (cost)`
    ) %>%
    
  # Make column of total costs with main treatment groupings.
  mutate(
    total_cost_new = rowSums(select(., 
      diagnostics_cost, prevent_tmnt_cost, restorations_cost, perio_cost, 
      exo_act_cost, endo_rct_cost, surg_oth_cost, prosth_cost), na.rm = TRUE)
  )
    
```


Compare total_cost_new to previous total_cost column made in Excel (sanity check).
```{r}
metadata <- metadata %>%
  mutate(
    total_cost = as.numeric(total_cost),
    total_cost_new = as.numeric(total_cost_new),
    match_status = if_else(total_cost_new == total_cost, "Match", "Mismatch"),
    difference = total_cost_new - total_cost # Optional: Calculate the difference for mismatches
  )


## Filter rows where there is a mismatch
mismatches <- metadata %>%
  filter(match_status == "Mismatch") %>%
  select(p_ID, total_cost, total_cost_new) %>%
  print()

```

All is okay, 295 and 39C come up due to differences in decimal places.

Format categorical variables.
```{r}
# Location
metadata <- metadata %>%
  mutate(
    remoteness_group = case_when(
      remoteness %in% c("Major Cities of Australia", "Inner Regional Australia") ~ "Metropolitan and Inner Regional",
      remoteness %in% c("Outer Regional Australia", "Remote Australia", "Very Remote Australia") ~ "Outer Regional, Remote, and Very Remote"
    )
  )

# Age
metadata$age_range <- ifelse(metadata$age < 18, "Under 18",
                             ifelse(metadata$age > 17 & metadata$age < 30, "18 to 29",
                                ifelse(metadata$age > 29 & metadata$age < 46, "30 to 45", 
                                    ifelse(metadata$age > 45 & metadata$age < 61, "46 to 60", 
                                           ifelse(metadata$age > 60, "Over 60", NA)))))

metadata$age_range <- factor(metadata$age_range, order = TRUE, 
                           levels = c("Under 18", "18 to 29", "30 to 45", "46 to 60", "Over 60"))

```


## 2. Format data into tables
Format Table 2.
```{r}
summary_table <- metadata %>%
  group_by(age_range) %>%
  summarize(
    Diagnostics = sum(diagnostics_cost, na.rm = TRUE),
    Preventive = sum(prevent_tmnt_cost, na.rm = TRUE),
    Restorations = sum(restorations_cost, na.rm = TRUE),
    Periodontics = sum(perio_cost, na.rm = TRUE),
    Extractions = sum(exo_act_cost, na.rm = TRUE),
    Endodontics = sum(endo_rct_cost, na.rm = TRUE),
    "Other surgical" = sum(surg_oth_cost, na.rm = TRUE),
    Prosthodontics = sum(prosth_cost, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = -age_range,
    names_to = "Treatment_Type",
    values_to = "Cost"
  ) %>%
  group_by(age_range) %>%
  mutate(
    Total_Cost = sum(Cost),
    Percentage = (Cost / Total_Cost) * 100
  ) %>%
  ungroup()


# Reshape data into wide format for the table
final_table <- summary_table %>%
  mutate(
    Cost_Percentage = paste0("$", formatC(Cost, format = "f", big.mark = ",", digits = 0), 
                             " (", formatC(Percentage, format = "f", digits = 1), "%)")
  ) %>%
  select(Treatment_Type, age_range, Cost_Percentage) %>%
  pivot_wider(
    names_from = age_range,
    values_from = Cost_Percentage,
    names_glue = "{age_range}"
  )


print(final_table)
```

Format Table 2 to include number of participants in each age range group in column headings.
```{r}
# Add counts for each age range group
age_counts <- metadata %>%
  count(age_range) %>%
  mutate(age_range_label = paste0(age_range, " (n = ", n, ")"))

summary_table <- summary_table %>%
  left_join(age_counts, by = "age_range") %>%
  mutate(age_range = age_range_label) %>%
  select(-n, -age_range_label)  # Remove extra columns after merging

final_table <- summary_table %>%
  mutate(
    Cost_Percentage = paste0("$", formatC(Cost, format = "f", big.mark = ",", digits = 0), 
                             " (", formatC(Percentage, format = "f", digits = 1), "%)")
  ) %>%
  select(Treatment_Type, age_range, Cost_Percentage) %>%
  pivot_wider(
    names_from = age_range,
    values_from = Cost_Percentage,
    names_glue = "{age_range}"
  )

print(final_table)
```

Format table to include n values for each age group.
```{r}
# Count participants in each age group
age_counts <- metadata %>%
  count(age_range) %>%
  mutate(age_range = paste0(age_range, " (n = ", n, ")")) %>%
  select(-n)  # Remove count column after merging

# Summarize treatment costs by age range
summary_table <- metadata %>%
  group_by(age_range) %>%
  summarize(
    Diagnostics = sum(diagnostics_cost, na.rm = TRUE),
    Preventive = sum(prevent_tmnt_cost, na.rm = TRUE),
    Restorations = sum(restorations_cost, na.rm = TRUE),
    Periodontics = sum(perio_cost, na.rm = TRUE),
    Extractions = sum(exo_act_cost, na.rm = TRUE),
    Endodontics = sum(endo_rct_cost, na.rm = TRUE),
    "Other surgical" = sum(surg_oth_cost, na.rm = TRUE),
    Prosthodontics = sum(prosth_cost, na.rm = TRUE)
  ) %>%
  left_join(age_counts, by = "age_range")  # Attach age group labels with participant counts

# Convert costs to percentage within each age group
summary_table <- summary_table %>%
  pivot_longer(
    cols = -age_range, 
    names_to = "Treatment_Type", 
    values_to = "Cost"
  ) %>%
  group_by(age_range) %>%
  mutate(
    Total_Cost = sum(Cost),
    Percentage = (Cost / Total_Cost) * 100
  ) %>%
  ungroup()

# Reshape data so age groups are rows and treatment types are columns
final_table <- summary_table %>%
  mutate(
    Cost_Percentage = paste0("$", formatC(Cost, format = "f", big.mark = ",", digits = 0), 
                             " (", formatC(Percentage, format = "f", digits = 1), "%)")
  ) %>%
  select(age_range, Treatment_Type, Cost_Percentage) %>%
  pivot_wider(
    names_from = Treatment_Type,
    values_from = Cost_Percentage
  )

# Print final table
print(final_table)
```


Summarise expenditure across total cohort.
```{r}
treatment_data <- metadata %>%
  summarize(
    Diagnostics = sum(diagnostics_cost, na.rm = TRUE),
    Preventive = sum(prevent_tmnt_cost, na.rm = TRUE),
    Restorations = sum(restorations_cost, na.rm = TRUE),
    Periodontics = sum(perio_cost, na.rm = TRUE),
    Extractions = sum(exo_act_cost, na.rm = TRUE),
    Endodontics = sum(endo_rct_cost, na.rm = TRUE),
    "Other surgical" = sum(surg_oth_cost, na.rm = TRUE),
    Prosthodontics = sum(prosth_cost, na.rm = TRUE))

print(treatment_data)
```

## 3.1 Figure 1A
Create pie chart plot for total expenditure.
```{r}
library(scales)
library(data.table)
library(ggrepel)
library(paletteer)

## Data transpose
treatment_data <- transpose(treatment_data, keep.names = "Treatment_type") 
treatment_data <- treatment_data %>%
  rename(Cost = V1)

## Create order of labels
treatment_data$Treatment_type <- factor(
  treatment_data$Treatment_type,
  levels = (c("Diagnostics",
              "Preventive",
             "Restorations", 
             "Extractions",
             "Endodontics",
             "Periodontics",
             "Prosthodontics",
             "Other surgical")
))

treatment_data <- treatment_data %>%
  arrange(Treatment_type)


## Calculate percentages, label positions, and make labels
treatment_data <- treatment_data %>%
  mutate(
    csum = rev(cumsum(rev(Cost))),
    pos = Cost / 2 + lead(csum, 1),
    pos = if_else(is.na(pos), Cost / 2, pos),
    Percentage = Cost / sum(Cost) * 100,
    Label = paste0("$", sprintf("%.1f", Cost / 1000), "k\n(", sprintf("%.1f", Percentage), "%)")
  )


## Reorder colour palette
#colours <- palettes_d$MetBrewer$Thomas
#colours
custom_colours <- c("#e1896e", 
                    "#c44d76", 
                    "#4457a5", 
                    "#13315f", 
                    "#b1a1cc", 
                    "#59386c", 
                    "#447861", 
                    "#7caf5c")



## Create pie chart
plot1 <- ggplot(treatment_data, aes(x = 2, y = Cost, fill = Treatment_type)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  #Internal labels
  geom_text(aes(label = ifelse(Percentage > 2, Label, "")),
            position = position_stack(vjust = 0.5),
            colour = "white",
            size = 2.8
  ) +
  #External labels for small slices
  geom_text_repel(data = treatment_data %>% filter(Percentage < 2),
                  aes(x = 2, y = pos, label = Label), 
                  size = 2.8,
                  nudge_x = 0.8,
                  segment.size = 0,
                  show.legend = FALSE
) +
  xlim(0.4, 3) +  #Creates the donut hole
  theme_void() +
  labs(fill = "Treatment type") +
  theme(
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    legend.position = "right"
  ) +
  scale_fill_manual(values = custom_colours) +
  #Centre label
    geom_text(
    x = 0, y = 0,
    label = "$290,436\n(100%)",
    size = 3,
    lineheight = 1.2,
    vjust = -0.5,
    hjust = 0.5
  )


plot1


ggsave(filename = "pie.tiff",
       height = 6, width = 8, dpi = 300, device='tiff')

```


## 3.2 Figure 1B
```{r}
library(ggplot2)

## Round percentages in summary table to 1 decimal place
summary_table <- summary_table %>%
  mutate(across(where(is.numeric), round, 1))

## Create levels for age groups and treatment types
summary_table$age_range <- factor(
  summary_table$age_range,
  levels = c("Under 18", 
             "18 to 29", 
             "30 to 45", 
             "46 to 60", 
             "Over 60")
)

## Create vector of n values (order must match the age_range factor levels)
n_values <- c("Under 18" = expression(atop(paste(bold("Under 18")), "(n = 9)")),
              "18 to 29" = expression(atop(paste(bold("18 to 29")), "(n = 28)")),
              "30 to 45" = expression(atop(paste(bold("30 to 45")), "(n = 51)")),
              "46 to 60" = expression(atop(paste(bold("46 to 60")), "(n = 57)")),
              "Over 60" = expression(atop(paste(bold("Over 60")), "(n = 28)")))


summary_table$Treatment_Type <- factor(
  summary_table$Treatment_Type,
  levels = (c("Diagnostics",
              "Preventive",
             "Restorations", 
             "Extractions",
             "Endodontics",
             "Periodontics",
             "Prosthodontics",
             "Other surgical")
))



## Create a stacked horizontal bar chart
plot2 <- ggplot(summary_table, aes(x = age_range, y = Percentage, fill = Treatment_Type)) +
  geom_bar(stat = "identity", colour = "white", width = 0.8) +
  #coord_flip() +  #For horizontal bar chart
  geom_text(aes(label = ifelse(Percentage >= 3, paste0(sprintf("%.1f", Percentage), "%"), "")),
  position = position_stack(vjust = 0.5),
  colour = "white",
  size = 5) +
  geom_text(aes(label = ifelse(Percentage < 3 & Percentage > 0, paste0(sprintf("%.1f", Percentage), "%"), "")),
  position = position_stack(vjust = 0.5),
  colour = "white",
  size = 2.5) +
  labs(
    x = "Age group (years)",
    y = "Percentage of total cost",
    fill = "Treatment type"
  ) +
  theme_bw() +
 scale_x_discrete(labels = n_values, expand = c(0, 0.45)) + # Remove vertical space above/below bars
 scale_y_continuous(labels = scales::percent_format(scale = 1), expand = expansion(mult = c(0.01, 0))) +  # Remove left/right padding
  theme(
    axis.title = element_text(face = "bold", size = 14),
    axis.title.x = element_text(vjust = -1.8),
    axis.text = element_text(size = 14, colour = "black"),
    axis.line.x = element_line(colour = "black"),
    axis.line.y = element_line(colour = "black"),
    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(size = 12),
    legend.position = "none",
    panel.border = element_blank(),
    panel.grid = element_blank(),
    panel.spacing = unit(0.1, "lines"),
    plot.margin = margin(5, 5, 10, 5)
    ) +
  scale_fill_manual(values = custom_colours)


plot2
```

Save plot for journal format.
```{r}
ggsave(filename = "bar.tiff",
       height = 10, width = 8, dpi = 300, device='tiff')
```



## 3.3 Figure 1 legend
```{r}
## Combine with patchwork.
library(patchwork)
library(cowplot)


## Make plot to extract legend from
plot_legend <- ggplot(treatment_data, aes(x = 2, y = Cost, fill = Treatment_type)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  theme_void() +
  labs(fill = "Treatment\ntype", ) +
  theme(legend.position = "bottom",
    legend.title = element_text(size = 12, hjust = 0.5),
    legend.text = element_text(size = 11),
    legend.key.size = unit(0.8, "cm")
  )  +
  scale_fill_manual(values = custom_colours)

plot_legend

## Find which grob the legend is contained in
#all_guides <- cowplot::get_plot_component(plot1, "guide-box", return_all = TRUE)
#all_guides

## Pull the legend from the third grob
#legend <- all_guides[[3]]
#plot_grid(legend)

```

Save legend.
```{r}
ggsave(filename = "legend.tiff",
       height = 10, width = 10, dpi = 300, device='tiff')
```

Full Figure 1 made by combining images in Powerpoint.