---
title: "Table S1"
author: "Gina Guzzo"
date: "17/01/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Load libraries and packages.

```{r, include = FALSE}
library(kableExtra)
library(rmarkdown)
set.seed(42)
library(qwraps2)
# define the markup language we are working in.
# options(qwraps2_markup = "latex") is also supported.
options(qwraps2_markup = "markdown")
library(readxl) 
library(dplyr)
library(tidyr)

knitr::opts_knit$set(root.dir = "/2025_dcs_financial_analysis/Analysis/")
```


Import data. 
Split cost columns that have more than one value into sub-categories, using commas as separators.

```{r, include = FALSE}
metadata <- read_excel("metadata/all_metadata_financial.xlsx")

metadata <- metadata %>%
  separate(`exam (qty, cost)`, into = c("exam_qty", "exam_cost"), sep = ", ", convert = TRUE) %>%
  separate(`bw (qty, cost)`, into = c("bw_qty", "bw_cost"), sep = ", ", convert = TRUE) %>%
  separate(`opg (qty, cost)`, into = c("opg_qty", "opg_cost"), sep = ", ", convert = TRUE) %>%
  separate(`fluor (qty, cost)`, into = c("fluor_qty", "fluor_cost"), sep = ", ", convert = TRUE) %>%
  separate(`scrp (qty, cost)`, into = c("scrp_qty", "scrp_cost"), sep = ", ", convert = TRUE) %>%
  separate(`perio (qty, cost)`, into = c("perio_qty", "perio_cost"), sep = ", ", convert = TRUE) %>%
  separate(`sealants (est qty, est cost, act qty, act cost)`, into = c("seal_est_qty", "seal_est_cost", "seal_act_qty", "seal_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`5_surf_resto (est qty, est cost, act qty, act cost)`, into = c("5_surf_resto_est_qty", "5_surf_resto_est_cost", "5_surf_resto_act_qty", "5_surf_resto_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`4_surf_resto (est qty, est cost, act qty, act cost)`, into = c("4_surf_resto_est_qty", "4_surf_resto_est_cost", "4_surf_resto_act_qty", "4_surf_resto_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`3_surf_resto (est qty, est cost, act qty, act cost)`, into = c("3_surf_resto_est_qty", "3_surf_resto_est_cost", "3_surf_resto_act_qty", "3_surf_resto_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`2_surf_resto (est qty, est cost, act qty, act cost)`, into = c("2_surf_resto_est_qty", "2_surf_resto_est_cost", "2_surf_resto_act_qty", "2_surf_resto_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`1_surf_resto (est qty, est cost, act qty, act cost)`, into = c("1_surf_resto_est_qty", "1_surf_resto_est_cost", "1_surf_resto_act_qty", "1_surf_resto_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`resto_temp (qty, cost)`, into = c("resto_tmp_qty", "resto_tmp_cost"), sep = ", ", convert = TRUE) %>%
  separate(`resto_other (qty, cost)`, into = c("resto_oth_qty", "resto_oth_cost"), sep = ", ", convert = TRUE) %>%
  separate(`exo (est qty, est cost, act qty, act cost)`, into = c("exo_est_qty", "exo_est_cost", "exo_act_qty", "exo_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`endo_rct (qty, cost)`, into = c("endo_rct_qty", "endo_rct_cost"), sep = ", ", convert = TRUE) %>%
  separate(`surg_other (qty, cost)`, into = c("surg_oth_qty", "surg_oth_cost"), sep = ", ", convert = TRUE) %>%
  separate(`occ_splint (est qty, est cost, act qty, act cost)`, into = c("occ_splint_est_qty", "occ_splint_est_cost", "occ_splint_act_qty", "occ_splint_act_cost"), sep = ", ", convert = TRUE) %>%
  separate(`ohi (qty, cost)`, into = c("ohi_qty", "ohi_cost"), sep = ", ", convert = TRUE) %>%
  separate(`odont (qty, cost)`, into = c("odont_qty", "odont_cost"), sep = ", ", convert = TRUE)
```


Clean up columns with costs to remove $ symbol.
```{r}
metadata <- metadata %>%
  mutate(
    exam_cost = as.numeric(gsub("\\$", "", exam_cost)),
    bw_cost = as.numeric(gsub("\\$", "", bw_cost)),
    opg_cost = as.numeric(gsub("\\$", "", opg_cost)),
    fluor_cost = as.numeric(gsub("\\$", "", fluor_cost)),
    scrp_cost = as.numeric(gsub("\\$", "", scrp_cost)),
    perio_cost = as.numeric(gsub("\\$", "", perio_cost)),
    seal_est_cost = as.numeric(gsub("\\$", "", seal_est_cost)),
    seal_act_cost = as.numeric(gsub("\\$", "", seal_act_cost)),
    `5_surf_resto_est_cost` = as.numeric(gsub("\\$", "", `5_surf_resto_est_cost`)),
    `5_surf_resto_act_cost` = as.numeric(gsub("\\$", "", `5_surf_resto_act_cost`)),
    `4_surf_resto_est_cost` = as.numeric(gsub("\\$", "", `4_surf_resto_est_cost`)),
    `4_surf_resto_act_cost` = as.numeric(gsub("\\$", "", `4_surf_resto_act_cost`)),
    `3_surf_resto_est_cost` = as.numeric(gsub("\\$", "", `3_surf_resto_est_cost`)),
    `3_surf_resto_act_cost` = as.numeric(gsub("\\$", "", `3_surf_resto_act_cost`)),
    `2_surf_resto_est_cost` = as.numeric(gsub("\\$", "", `2_surf_resto_est_cost`)),
    `2_surf_resto_act_cost` = as.numeric(gsub("\\$", "", `2_surf_resto_act_cost`)),
    `1_surf_resto_est_cost` = as.numeric(gsub("\\$", "", `1_surf_resto_est_cost`)),
    `1_surf_resto_act_cost` = as.numeric(gsub("\\$", "", `1_surf_resto_act_cost`)),
    resto_tmp_cost = as.numeric(gsub("\\$", "", resto_tmp_cost)),
    resto_oth_cost = as.numeric(gsub("\\$", "", resto_oth_cost)),
    exo_est_cost = as.numeric(gsub("\\$", "", exo_est_cost)),
    exo_act_cost = as.numeric(gsub("\\$", "", exo_act_cost)),
    endo_rct_cost = as.numeric(gsub("\\$", "", endo_rct_cost)),
    surg_oth_cost = as.numeric(gsub("\\$", "", surg_oth_cost)),
    occ_splint_est_cost = as.numeric(gsub("\\$", "", occ_splint_est_cost)),
    occ_splint_act_cost = as.numeric(gsub("\\$", "", occ_splint_act_cost)),
    ohi_cost = as.numeric(gsub("\\$", "", ohi_cost)),
    odont_cost = as.numeric(gsub("\\$", "", odont_cost)),
    `prosth (cost)` = as.numeric(gsub("\\$", "", `prosth (cost)`))
    
  )
```


Make broader categories for treatment types by summing columns.
```{r}
metadata <- metadata %>%
  mutate(
    # Sum diagnostic costs
    diagnostics_cost = rowSums(select(., exam_cost, bw_cost, opg_cost), na.rm = TRUE),
    
    # Sum preventative treatment costs
    prevent_tmnt_cost = rowSums(select(., 
      fluor_cost, scrp_cost, seal_act_cost, ohi_cost, odont_cost, occ_splint_act_cost), na.rm = TRUE),
    
    # Sum restoration costs
    restorations_cost = rowSums(select(., 
      `5_surf_resto_act_cost`, `4_surf_resto_act_cost`, 
      `3_surf_resto_act_cost`, `2_surf_resto_act_cost`, 
      `1_surf_resto_act_cost`, resto_tmp_cost, resto_oth_cost), na.rm = TRUE),
    
    # Sum prosthodontic costs
    prosth_cost = `prosth (cost)`
    ) %>%
    
  # Make column of total costs with main treatment groupings.
  mutate(
    total_cost_new = rowSums(select(., 
      diagnostics_cost, prevent_tmnt_cost, restorations_cost, perio_cost, 
      exo_act_cost, endo_rct_cost, surg_oth_cost, prosth_cost), na.rm = TRUE)
  )
    
```


Compare total_cost_new to previous total_cost column made in Excel.
```{r}
metadata <- metadata %>%
  mutate(
    total_cost = as.numeric(total_cost),
    total_cost_new = as.numeric(total_cost_new),
    match_status = if_else(total_cost_new == total_cost, "Match", "Mismatch"),
    difference = total_cost_new - total_cost # Optional: Calculate the difference for mismatches
  )


# Filter rows where there is a mismatch
mismatches <- metadata %>%
  filter(match_status == "Mismatch") %>%
  select(p_ID, total_cost, total_cost_new) %>%
  print()

```

All is okay, 295 comes up due to differences in decimal length.

Format categorical variables.
```{r}
# Location
metadata <- metadata %>%
  mutate(
    remoteness_group = case_when(
      remoteness %in% c("Major Cities of Australia", "Inner Regional Australia") ~ "Metropolitan and Inner Regional",
      remoteness %in% c("Outer Regional Australia", "Remote Australia", "Very Remote Australia") ~ "Outer Regional, Remote, and Very Remote"
    )
  )

# Age
metadata$age_range <- ifelse(metadata$age < 18, "Under 18",
                             ifelse(metadata$age > 17 & metadata$age < 30, "18-29",
                                ifelse(metadata$age > 29 & metadata$age < 46, "30-45", 
                                    ifelse(metadata$age > 45 & metadata$age < 61, "46-60", 
                                           ifelse(metadata$age > 60, "Over 60", NA)))))

metadata$age_range <- factor(metadata$age_range, order = TRUE, 
                           levels = c("Under 18", "18-29", "30-45", "46-60", "Over 60"))

```



Format Table S1.
```{r}
summary_table <- metadata %>%
  group_by(remoteness_group) %>%
  summarize(
    Diagnostics = sum(diagnostics_cost, na.rm = TRUE),
    Preventative = sum(prevent_tmnt_cost, na.rm = TRUE),
    Restorations = sum(restorations_cost, na.rm = TRUE),
    Periodontics = sum(perio_cost, na.rm = TRUE),
    Extractions = sum(exo_act_cost, na.rm = TRUE),
    Endodontics = sum(endo_rct_cost, na.rm = TRUE),
    Other_Surgical = sum(surg_oth_cost, na.rm = TRUE),
    Prosthodontics = sum(prosth_cost, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = -remoteness_group,
    names_to = "Treatment_Type",
    values_to = "Cost"
  ) %>%
  group_by(remoteness_group) %>%
  mutate(
    Total_Cost = sum(Cost),
    Percentage = (Cost / Total_Cost) * 100
  ) %>%
  ungroup()


# Reshape data into wide format for the table
final_table <- summary_table %>%
  mutate(
    Cost_Percentage = paste0("$", formatC(Cost, format = "f", big.mark = ",", digits = 0), 
                             " (", formatC(Percentage, format = "f", digits = 1), "%)")
  ) %>%
  select(Treatment_Type, remoteness_group, Cost_Percentage) %>%
  pivot_wider(
    names_from = remoteness_group,
    values_from = Cost_Percentage,
    names_glue = "{remoteness_group}"
  )


print(final_table)
```

Format Table S1 to include number of participants in each location group in column headings.
```{r}
# Add counts for each age range group
location_counts <- metadata %>%
  count(remoteness_group) %>%
  mutate(remoteness_group_label = paste0(remoteness_group, " (n = ", n, ")"))

summary_table <- summary_table %>%
  left_join(location_counts, by = "remoteness_group") %>%
  mutate(remoteness_group = remoteness_group_label) %>%
  select(-n, -remoteness_group_label)  # Remove extra columns after merging

final_table <- summary_table %>%
  mutate(
    Cost_Percentage = paste0("$", formatC(Cost, format = "f", big.mark = ",", digits = 0), 
                             " (", formatC(Percentage, format = "f", digits = 1), "%)")
  ) %>%
  select(Treatment_Type, remoteness_group, Cost_Percentage) %>%
  pivot_wider(
    names_from = remoteness_group,
    values_from = Cost_Percentage,
    names_glue = "{remoteness_group}"
  )

print(final_table)
```

Calculate median age for each location group.
```{r}
# Calculate median and IQR of age by remoteness group
age_summary <- metadata %>%
  group_by(remoteness_group) %>%
  summarize(
    N = n(),
    Median_Age = median(age, na.rm = TRUE),
    Q1 = quantile(age, 0.25, na.rm = TRUE),
    Q3 = quantile(age, 0.75, na.rm = TRUE)
  ) %>%
  mutate(
    IQR = paste0(Median_Age, " (", Q1, "–", Q3, ")")
  ) %>%
  select(remoteness_group, N, IQR)


# Display the result
knitr::kable(age_summary, caption = "Median Age and IQR by Location Group")
```

